Live coding

### 2019-01-23 wip-live-programming/paper.tex
Updating correctness proofs

\theoremstyle{definition}
\newtheorem{definition}[dfc]{Definition}

@@

\newcommand{\bndclr}[1]{\textcolor{kvdclr}{#1}}
\newcommand{\bkndclr}[1]{\textcolor{prepclr}{#1}}
\newcommand{\blblclr}[1]{\textcolor{numclr}{#1}}
\newcommand{\bnd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\bknd}[1]{\textnormal{\textcolor{prepclr}{\sffamily #1}}}
\newcommand{\blbl}[1]{\textnormal{\textcolor{numclr}{\sffamily #1}}}
\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}

@@

Increasing number of programming environments aim to provide the same live experience for more
standard programming languages, but doing this is not easy. Fully recomputing the whole program after
every single change is inefficient and calculating how a change in source code changes the result
is extremely hard when the editor allows arbitrary manipulation of program text. For example,
consider the following simple program that gets the release years of 10 most expensive movies
in a data set \ident{movies}:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{top} =\ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\num{10})\\
\quad .\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))}\\
\end{array}
\end{equation*}
%
A live coding environment shows a preview of the list of dates. Next, assume that the programmer
modifies the code by making the constant $\num{10}$ a variable and changing the date format to see
the full date:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{count} = \num{10}\\
\kvd{let}~\ident{top} = \ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\ident{count})\\
\quad \ident{.map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))}\\
\end{array}
\end{equation*}
%
Ideally, the live coding environment should understand the change, reuse a cached result of the
first two transformations (sorting and taking 10 elements) and only evaluate the last \ident{map}
to differently format the release dates of already computed top 10 movies.

@@

\newpage
\begin{figure}
\raggedright
{\small\sffamily Programs, commands, expressions, arguments and values}
\begin{equation*}
\hspace{-2em}
\begin{array}{lcl}
p &\narrow{=}& c_1; \ldots; c_n\\
c &\narrow{=}& \kvd{let}~x = e \lsep e\\
\end{array}
\qquad
\begin{array}{lcl}
e &\narrow{=}& o \lsep x \\
  &\narrow{|}& e.m(a, \ldots, a)
\end{array}
\qquad
\begin{array}{lcl}
a &\narrow{=}& e \lsep \lambda x\rightarrow e\\
v &\narrow{=}& o \lsep \lambda x\rightarrow e\\
\end{array}
\end{equation*}

\vspace{0.5em}
{\small\sffamily Evaluation contexts of expressions}
\begin{equation*}
\hspace{-2em}
\begin{array}{lcl}
C_e[-] &=& -.m(a_1, \ldots, a_n) ~\lsep~ -.m(v_1, \ldots, v_{m}, -, a_1,\ldots , a_n) ~\lsep~ -\\
C_c[-] &=& \kvd{let}~x = C_e[-] \hspace{1.25em}\lsep~ C_e[-]\\
C_p[-] &=& o_1;\; \ldots;\; o_{k};\; C_c[-];\; c_{1};\; \ldots;\; c_n\\
\end{array}
\end{equation*}

\vspace{0.5em}
{\small\sffamily Let elimination and member reduction}
\begin{equation*}
\hspace{-2em}
\begin{array}{ll}
\hspace{-0.5em}\begin{array}{l}
o_1;\; \ldots;\; o_{k};\; \kvd{let}~x=o;\; c_{1};\; \ldots; c_n \rightsquigarrow\\
\qquad  o_1;\; \ldots;\; o_{k};\; c_{1}[x\leftarrow o];\; \ldots; c_n[x\leftarrow o]\end{array} &
\quad\text{\small\sffamily(let)}\\[1.2em]
{o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'}\implies
  {C_p[o.m(v_1, \ldots, v_n)] \rightsquigarrow C_p[o']} &
\quad\text{\small\sffamily (external)}
\end{array}
\end{equation*}

@@

A program $p$ in the data exploration calculus consists of a sequence of commands $c$. A command
can be either let binding or an expression. Let bindings define variables $x$ that can be used in
subsequent commands. The calculus supports lambda functions, but they
can only appear as arguments in method calls. An argument $a$ can thus be a lambda
function or an expression.\footnote{Similar but weaker restrictions on the use of lambda functions
exist in other languages. To guide type inference, lambda functions in C\# can appear as either
method arguments or assigned to an explicitly typed variable.}

\subsection{Operational semantics}
The data exploration calculus is a call-by-value language. We model the evaluation as a small-step
reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of
objects $o_1;\; \ldots;\; o_n$ which can be displayed as intermediate results of the data analysis. The
operational semantics is parameterized by a relation $\rightsquigarrow_\epsilon$ that models
the functionality of the external libraries used with the calculus and defines reduction rules
for member accesses. The relation has the following form:
%
\begin{equation*}
o_1.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o_2
\end{equation*}
%
Figure~\ref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
and evaluation contexts; $C_e$ specifies left-to-right evaluation of arguments of a method call,
$C_c$ specifies evluation of a command and $C_p$ defines top-to-bottom evaluation of a program.
The rule {\small\sffamily (external)} applies reductions provided by an external library in a
call-by-value order and {\small\sffamily (let)} substitutes a value of an evaluated variable in
all subsequent commands.

@@

TBD: Some adaptation of the stuff below, as needed for the proofs.

@@

% TODO: Adapt as needed
%
% \subsection{Equational theory}
% In The Gamma language, computations are expressed using
% member access, written as $e.m(e_1,\ldots, e_n)$. In this paper, we do not define how member
% access evaluates. This has been done elsewhere \cite{gamma}, but more importantly, the evaluation of
% previews does not rely on the exact specifics of the evaluation, provided that the language
% satisfies certain basic conditions. The following definitions provides the necessary structure
% for discussing correctness of previews.
%
% Partial evaluation may reduce an expression under $\lambda$-abstraction. We do not require that
% the reduction of the host language does this. Instead, we define an extended reduction relation
% and use that in the proofs. The host language only needs to compose well with such extended reduction
% as captured by the \emph{compositionality} property below. We also require that the language allows
% elimination of let bindings.
%
% \begin{definition}[Host language]
% \label{def:host}
% Given a relation on expressions $e_1 \rightsquigarrow e_2$ that models small-step evaluation, we define:
%
% \begin{itemize}
% \item[--]
%  A \emph{preview evaluation context} (also referred to as \emph{context}):
%
% \vspace{-1.25em}
% \begin{equation*}
% \begin{array}{lcl}
% C[-] &\hspace{-1em}=\hspace{-1em}& \kvd{let}~x = -~\kvd{in}~e ~|~ \kvd{let}~x = e~\kvd{in}~- ~|~ \lambda x\rightarrow - \\
%      &&  e_0.m(e_1, \ldots, e_{k-1}, -, e_{k+1}, \ldots e_n)
% \end{array}
% \end{equation*}
%
% \item[--]
% An extended reduction relation $\rightsquigarrow_\beta$ such that, for any context $C$,
% $C[e_1] \rightsquigarrow_\beta C[e_2]$ whenever $e_1 \rightsquigarrow e_2$.
%
% \item[--]
% Let elimination $\rightsquigarrow_\ident{let}$ such that, using  capture-avoiding substitution,
%   $C[\kvd{let}~x=e_1~\kvd{in}~e_2] \rightsquigarrow_\ident{let} C[e_2[x\leftarrow e_1]]$
% \end{itemize}
%
% \noindent
% We say that $\rightsquigarrow$ is a suitable \emph{host language reduction} if:
% \begin{itemize}
% \item[--] It satisfies the \emph{compositionality} property, that is if $e \rightsquigarrow e'$ and
% $C[e]\rightsquigarrow_\beta e''$ then also $C[e']\rightsquigarrow_\beta e''$.
%
% \item[--] Let elimination does not affect the result, i.e.~if $e \rightsquigarrow_{\ident{let}} e'$
% and $e'\rightsquigarrow_\beta e''$ then also $e\rightsquigarrow_\beta e''$
% \end{itemize}
% \end{definition}
%
% \noindent
% The host language in The Gamma is a simple call-by-value functional language without side-effects,
% and so it satisfies both compositionality and allows let bindings to be eliminated, although the
% latter affects the performance. The mechanism for preview evaluation presented here would also
% work for call-by-name languages, but it would suffer from the expected difficulties in the
% presence of side-effects or non-determinism.
%
% \newpage

@@

\section{Formalising live coding infrastructure}
\label{sec:formal}

@@

\begin{figure*}[t]
\begin{equation*}
\hspace{-3.5em}
\begin{array}{ll}
\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n) &(1)~~\\
\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta}(e_i)
~\textnormal{and}~\bndclr{v} = \Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))])\\
\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\}
\\[1em]
\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)&(2)\\
\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta}(e_i)
~\textnormal{and}~\Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))])\!\downarrow\\
\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}
~\textnormal{and}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
\\[1em]
\ident{bind-expr}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(3)\\
\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}
~\textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind-expr}_{\Gamma_1, \Delta}(e)
~\textnormal{and}~\bndclr{v} = \Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\\
\quad \textnormal{let}~e = (\bndclr{v}, \bndclr{v_0}, \blbl{body})
\\[1em]
\ident{bind-expr}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(4)\\
\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}
~\textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind-expr}_{\Gamma_1, \Delta}(e)
~\textnormal{and}~\Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\!\downarrow\\
\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s), s~\textnormal{fresh}
~\textnormal{and}~e = (\bndclr{v},\bndclr{v_0},\blbl{body})
\\[1em]
\ident{bind-expr}_{\Gamma, \Delta}(o) = \bnd{val}(o), (\{ \bnd{val}(o) \}, \emptyset) }&(5)
\\[1em]
\ident{bind-expr}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)~\textnormal{when}~\bndclr{v} = \Gamma(x)&(6)
\end{array}
\end{equation*}
\caption{Binding rules that define a construction of a dependency graph for an expression.}
\label{fig:binding-rules-expr}
\end{figure*}

@@

\begin{figure*}[t]
\begin{equation*}
\hspace{-3.5em}
\begin{array}{ll}
\ident{bind-prog}_{\Gamma, \Delta}(\kvd{let}~x=e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v}\} \cup V \cup V_1, E \cup E_1)
  \hspace{9.35em}&(7)\\
\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind-expr}_{\Gamma,\Delta}(e_1)~\textnormal{and}~\Gamma_1 = \Gamma \cup \{(x,\bndclr{v_1})\} \\
\quad \textnormal{and}~\bndclr{v_2}; \ldots; \bndclr{v_n}, (V, E) = \ident{bind-prog}_{\Gamma_1, \Delta}(c_2; \ldots; c_n)
\\[1em]
\ident{bind-prog}_{\Gamma, \Delta}(e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v}\} \cup V \cup V_1, E \cup E_1)&(8)\\
\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind-expr}_{\Gamma,\Delta}(e)
~\textnormal{and}~\bndclr{v_2}; \ldots; \bndclr{v_n}, (V, E) = \ident{bind-prog}_{\Gamma_1, \Delta}(c_2; \ldots; c_n)
\\[1em]
\ident{bind-prog}_{\Gamma, \Delta}([]) = [], (\emptyset, \emptyset)&(9)
\end{array}
\end{equation*}
\caption{Binding rules that define a construction of a dependency graph for a program.}
\label{fig:binding-rules-prog}
\end{figure*}

@@

The mechanism for constructing dependency graphs defined in Section~\ref{sec:formal} makes it
possible to provide live previews when editing code without recomputing the whole program
each time the source code changes.

The nodes in the dependency graph correspond to individual operations that will be performed
when running the program. When the dependencies of an operation do not change while editing
code, the subsequent dependency graph will reuse a node used to represent the operation.

Our live editor keeps a map from graph nodes to live previews, so a new preview only needs to be
computed when a new node appears in the dependency graph (and the user moves the cursor to a
code location that corresponds to the node). This section describes how previews are evaluated.

@@

\begin{figure*}
\begin{equation*}
\qquad\qquad\quad\begin{array}{l}
\inference[(lift-expr)~]
  {\bndclr{v} \Downarrow \llbracket e \rrbracket_\Gamma}
  {\bndclr{v} \Downarrow_{\ident{lift}} \llbracket e \rrbracket_\Gamma}
\\
\\
\inference[(lift-prev)~]
  {\bndclr{v} \Downarrow p}
  {\bndclr{v} \Downarrow_{\ident{lift}} \llbracket p \rrbracket_\emptyset}
\\
\\
\inference[(val)~]
  {~}
  {\bnd{val}(n) \Downarrow n }
\\
\\
\inference[(var)~]
  {~}
  {\bnd{var}(x) \Downarrow \llbracket x \rrbracket_x}
  \\[0.5em]~
\end{array}
\qquad
\begin{array}{l}
\inference[(fun-val)~]
  { (\bnd{fun}(x, s), \bndclr{v}, \blbl{body}) \in E & \bndclr{v} \Downarrow p }
  { \bnd{fun}(x, s) \Downarrow \lambda x\rightarrow p }
\\
\\
\inference[(fun-bind)~]
  { (\bnd{fun}(x, s), \bndclr{v}, \blbl{body}) \in E & \bndclr{v} \Downarrow \llbracket e \rrbracket_{x} }
  { \bnd{fun}(x, s) \Downarrow \lambda x\rightarrow e }
\\
\\
\inference[(fun-expr)~]
  { (\bnd{fun}(x, s), \bndclr{v}, \blbl{body}) \in E & \bndclr{v} \Downarrow \llbracket e \rrbracket_{x,\Gamma} }
  { \bnd{fun}(x, s) \Downarrow \llbracket \lambda x\rightarrow e \rrbracket_\Gamma }
\\[0.5em]~
\end{array}
\end{equation*}
\begin{equation*}
\qquad\qquad\begin{array}{l}
\inference[(mem-val)~]
  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E & \bndclr{v_i} \Downarrow p_i
    & p_0.m(p_1, \ldots, p_k) \rightsquigarrow p }
  { \bnd{mem}(m, s) \Downarrow p }
\\
\\
\inference[(mem-expr)~]
  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E & \exists j\in\{0\ldots k\}.\bndclr{v_j}\diagup\hspace{-1em}\Downarrow p_j
   & \bndclr{v_i} \Downarrow_{\ident{lift}} \llbracket e_i \rrbracket_{\Gamma_i}  }
  { \bnd{mem}(m, s) \Downarrow \llbracket e_0.m(e_1, \ldots, e_k) \rrbracket_{\Gamma_0,\ldots,\Gamma k} }
\\
\end{array}
\end{equation*}

\caption{Rules that define evaluation of previews over a dependency graph for an expression}
\label{fig:eval}
\end{figure*}

@@

>>>>

\theoremstyle{definition}
\newtheorem{axiom}[dfc]{Axiom}
\newtheorem{definition}[dfc]{Definition}

@@

\newcommand{\bndclr}[1]{\textcolor{kvdclr}{#1}}
\newcommand{\bkndclr}[1]{\textcolor{prepclr}{#1}}
\newcommand{\blblclr}[1]{\textcolor{numclr}{#1}}
\newcommand{\bnd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\bknd}[1]{\textnormal{\textcolor{prepclr}{\sffamily #1}}}
\newcommand{\blbl}[1]{\textnormal{\textcolor{numclr}{\sffamily #1}}}
\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
\newcommand{\rname}[1]{{\sffamily\small(#1)}}

@@

Increasing number of programming environments aim to provide the same live experience for more
standard programming languages, but doing this is not easy. Fully recomputing the whole program after
every single change is inefficient and calculating how a change in source code changes the result
is extremely hard when the editor allows arbitrary manipulation of program text. For example,
consider the following simple program that gets the release years of 10 most expensive movies
in a data set \ident{movies}:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{top} =\ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\num{10})\\
\quad .\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))\\
\end{array}
\end{equation*}
%
A live coding environment shows a preview of the list of dates. Next, assume that the programmer
modifies the code by making the constant $\num{10}$ a variable and changing the date format to see
the full date:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{count} = \num{10}\\
\kvd{let}~\ident{top} = \ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\ident{count})\\
\quad \ident{.map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))\\
\end{array}
\end{equation*}
%
Ideally, the live coding environment should understand the change, reuse a cached result of the
first two transformations (sorting and taking 10 elements) and only evaluate the last \ident{map}
to differently format the release dates of already computed top 10 movies.

@@

\newpage
\begin{figure}
\raggedright
{\small\sffamily Programs, commands, terms, expressions and values}
\begin{equation*}
\hspace{-2em}
\begin{array}{lcl}
p &\narrow{=}& c_1; \ldots; c_n\\
c &\narrow{=}& \kvd{let}~x = t \lsep t\\
\end{array}
\qquad
\begin{array}{lcl}
t &\narrow{=}& o \lsep x \\
  &\narrow{|}& t.m(e, \ldots, e)
\end{array}
\qquad
\begin{array}{lcl}
e &\narrow{=}& t \lsep \lambda x\rightarrow e\\
v &\narrow{=}& o \lsep \lambda x\rightarrow e\\
\end{array}
\end{equation*}

\vspace{0.5em}
{\small\sffamily Evaluation contexts of expressions}
\begin{equation*}
\hspace{-2em}
\begin{array}{lcl}
C_e[-] &=& C_e[-].m(e_1, \ldots, e_n) ~\lsep~ o.m(v_1, \ldots, v_{m}, C_e[-], e_1,\ldots , e_n) ~\lsep~ -\\
C_c[-] &=& \kvd{let}~x = C_e[-] ~\lsep~ C_e[-]\\
C_p[-] &=& o_1;\; \ldots;\; o_{k};\; C_c[-];\; c_{1};\; \ldots;\; c_n\\
\end{array}
\end{equation*}

\vspace{0.5em}
{\small\sffamily Let elimination and member reduction}
\begin{equation*}
\hspace{-2em}
\begin{array}{ll}
\hspace{-0.5em}\begin{array}{l}
o_1;\; \ldots;\; o_{k};\; \kvd{let}~x=o;\; c_{1};\; \ldots; c_n \rightsquigarrow\\
\qquad  o_1;\; \ldots;\; o_{k};\; o; c_{1}[x\leftarrow o];\; \ldots; c_n[x\leftarrow o]\end{array} &
\quad\text{\rname{let}}\\[1.2em]
{o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'}\implies
  {C_p[o.m(v_1, \ldots, v_n)] \rightsquigarrow C_p[o']} &
\quad\text{\rname{external}}
\end{array}
\end{equation*}

@@

A program $p$ in the data exploration calculus consists of a sequence of commands $c$. A command
can be either let binding or a term. Let bindings define variables $x$ that can be used in
subsequent commands. The calculus supports lambda functions, but they
can only appear as arguments in method calls. A term $t$ can be a value, variable or a
member access, while an expression $e$, which can appear as an argument in member access,
can be a lambda function or a term.\footnote{Similar but weaker restrictions on the use of lambda functions
exist in other languages. To guide type inference, lambda functions in C\# can appear as either
method arguments or assigned to an explicitly typed variable.}

\subsection{Operational semantics}
\label{sec:calculus-semantics}

The data exploration calculus is a call-by-value language. We model the evaluation as a small-step
reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of
objects $o_1;\; \ldots;\; o_n$ which can be displayed as intermediate results of the data analysis. The
operational semantics is parameterized by a relation $\rightsquigarrow_\epsilon$ that models
the functionality of the external libraries used with the calculus and defines reduction rules
for member accesses. The relation has the following form:
%
\begin{equation*}
o_1.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o_2
\end{equation*}
%
Figure~\ref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
and evaluation contexts; $C_e$ specifies left-to-right evaluation of arguments of a method call,
$C_c$ specifies evluation of a command and $C_p$ defines top-to-bottom evaluation of a program.
The rule \rname{external} applies reductions provided by an external library in a
call-by-value order and \rname{let} substitutes a value of an evaluated variable in
all subsequent commands. Note that \rname{let} leaves the value of the variable
in the resulting list of commands.

@@

DEFINE $\rightsquigarrow^{*}$ as a reflexive, transitive closure of $\rightsquigarrow$

DEFINE $\rightsquigarrow_\ident{let}$ (on a program) - CBN subsitution

\begin{definition}[External library]
\label{def:external}

Satsifies \emph{compositionality} such that:

Given $e_0, e_1, \ldots, e_n$ and $e'_0, e'_1, \ldots, e'_n$ and $m$ such that
$e_0.m(e_1, \ldots, e_n) \rightsquigarrow^{*} o$ and
$e'_0.m(e'_1, \ldots, e'_n) \rightsquigarrow^{*} o'$ then
if for any contexts
$C_0, C_1, \ldots, C_n$ it holds that if $C_i[e_i] \rightsquigarrow^{*} o_i$ and $C_i[e'_i] \rightsquigarrow^{*} o_i$ for some $o_i$
then $o = o'$.
\end{definition}

TBD: LET above should leave the evaluated object in the list of commmands

LET ELIMINATION DOES NOT CHANGE RESULTS

Let elimination $\rightsquigarrow_\ident{let}$ such that, using  capture-avoiding substitution,
$C[\kvd{let}~x=e_1~\kvd{in}~e_2] \rightsquigarrow_\ident{let} C[e_2[x\leftarrow e_1]]$

Let elimination does not affect the result, i.e.~if $e \rightsquigarrow_{\ident{let}} e'$
and $e'\rightsquigarrow_\beta e''$ then also $e\rightsquigarrow_\beta e''$

\begin{lem}[Let elimination for a program]
\label{thm:let-lang-elimination}
does not matter
\end{lem}

@@

@@

\section{Formalising live coding environment}
\label{sec:formal}

@@

\begin{figure*}[t]
\begin{equation*}
\hspace{-3.5em}
\begin{array}{ll}
\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n) &(1)~~\\
\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta}(e_i)
~\textnormal{and}~\bndclr{v} = \Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))])\\
\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\}
\\[1em]
\ident{bind-expr}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)&(2)\\
\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind-expr}_{\Gamma, \Delta}(e_i)
~\textnormal{and}~\Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))])\!\downarrow\\
\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}
~\textnormal{and}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
\\[1em]
\ident{bind-expr}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(3)\\
\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}
~\textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind-expr}_{\Gamma_1, \Delta}(e)
~\textnormal{and}~\bndclr{v} = \Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\\
\quad \textnormal{let}~e = (\bndclr{v}, \bndclr{v_0}, \blbl{body})
\\[1em]
\ident{bind-expr}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(4)\\
\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}
~\textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind-expr}_{\Gamma_1, \Delta}(e)
~\textnormal{and}~\Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\!\downarrow\\
\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s), s~\textnormal{fresh}
~\textnormal{and}~e = (\bndclr{v},\bndclr{v_0},\blbl{body})
\\[1em]
\ident{bind-expr}_{\Gamma, \Delta}(o) = \bnd{val}(o), (\{ \bnd{val}(o) \}, \emptyset) &(5)
\\[1em]
\ident{bind-expr}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)~\textnormal{when}~\bndclr{v} = \Gamma(x)&(6)
\end{array}
\end{equation*}
\caption{Binding rules that define a construction of a dependency graph for an expression.}
\label{fig:binding-rules-expr}
\end{figure*}

@@

\begin{figure*}[t]
\begin{equation*}
\hspace{-3.5em}
\begin{array}{ll}
\ident{bind-prog}_{\Gamma, \Delta}(\kvd{let}~x=e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v_1}\} \cup V \cup V_1, E \cup E_1)
  \hspace{9.35em}&(7)\\
\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind-expr}_{\Gamma,\Delta}(e_1)~\textnormal{and}~\Gamma_1 = \Gamma \cup \{(x,\bndclr{v_1})\} \\
\quad \textnormal{and}~\bndclr{v_2}; \ldots; \bndclr{v_n}, (V, E) = \ident{bind-prog}_{\Gamma_1, \Delta}(c_2; \ldots; c_n)
\\[1em]
\ident{bind-prog}_{\Gamma, \Delta}(e; c_2; \ldots; c_n) = \bndclr{v_1};\ldots;\bndclr{v_n}, (\{\bndclr{v_1}\} \cup V \cup V_1, E \cup E_1)&(8)\\
\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind-expr}_{\Gamma,\Delta}(e)
~\textnormal{and}~\bndclr{v_2}; \ldots; \bndclr{v_n}, (V, E) = \ident{bind-prog}_{\Gamma_1, \Delta}(c_2; \ldots; c_n)
\\[1em]
\ident{bind-prog}_{\Gamma, \Delta}([]) = [], (\emptyset, \emptyset)&(9)
\end{array}
\end{equation*}
\caption{Binding rules that define a construction of a dependency graph for a program.}
\label{fig:binding-rules-prog}
\end{figure*}

@@

The binding process described in the previous section constructs a dependency graph after code
changes. The nodes in the dependency graph correspond to individual operations that will be performed
when running the program. When evaluating a preview, we attach (partial) results to
nodes of the graph. Since the binding process reuses nodes when their dependencies do not change,
previews for expressions (or sub-expressions) of a program can be reused when updating a preview.

In this section, we describe how previews are evaluated. The evaluation is done over the dependency
graph, rather than over the structure of the program expressions as in the operational semantics
given in Section~\ref{sec:calculus-semantics}. We analyse the preview evaluation formally in
Section~\ref{sec:properties} and show that the resulting previews are the same as the result we
would get by directly evaluating the code and, also, that no recomputation occurs when code is
edited in certain ways.

@@

\subsection{Previews and delayed previews}

Programs in the data exploration calculus consist of sequence of commands. Those can always be
evaluated to a value with a preview that can be displayed to the user. However, we also support
previews for all sub-expressions. This can be problematic if the current sub-expression is inside
the body of a function. For example:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{top} = \ident{movies}.\ident{take}(\num{10}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased}().\ident{format}(\str{dd-mm-yyyy}))
\end{array}
\end{equation*}
%
Here, we can directly evaluate sub-expressions \ident{movies} and $\ident{movies}.\ident{take}(\num{10})$,
but not $x.\ident{getReleased}()$ and $x.\ident{getReleased}().\ident{format}(\str{dd-mm-yyyy})$
because they contain a free variable $x$. Our preview evaluation algorithm addresses this by
producing two kinds of previews. A \emph{fully evaluated preview} is just a value, while
a \emph{delayed preview} is a partially evalauted expression with free variables:

\begin{equation*}
\begin{array}{rcll}
p&=&o \lsep  \lambda x\rightarrow e&\quad(\textit{Fully evaluated previews})\\
d&=&p \lsep  \llbracket e \rrbracket_\Gamma&\quad(\textit{Evaluated and delayed previews})\\
\end{array}
\end{equation*}
%
A fully evaluated preview $p$ can be either a primitive object or a function value with no free
variables. A possibly delayed preview $d$ can be either an evaluated preview $p$ or an expression
$e$ that requires variables $\Gamma$. For simplicity, we use an untyped language and so $\Gamma$
is just a list of variables $x_1, \ldots, x_n$.

A delayed preview is not necessarily the body of a lambda function as it appears in the source
code. We partially evaluate sub-expressions of the body that do not have free variables or that
have free variables bound by an earlier let binding.

Our implementation does not currently display
delayed previews to the user, but there is a number of possible approaches for doing that.
Most interestingly, since lambda functions always appear as arguments of a member access, we could
force the evaluation of the surrounding expression, capture (a number of) values passed as inputs
to the lambda function and display a preview based on those. Finally, in Section~\ref{sec:design},
we also consider a more speculative design for an abstraction mechanism that supports live previews
that could, in some cases, replace lambda function.

@@

\section{TODO}

TBD: In section about dependency graphs, include a subsection discussing how this relates
to incremental computation (just like "Semantics of delayed previews" links that to CMTT etc.)
